\documentclass[twoside,a4paper]{book}
\usepackage{fricas}
\title{The \Language{} System for Computer Mathematics}

\begin{document}
\maketitle
\pagenumbering{roman}%
\setcounter{page}{0}


\section*{General note}
This is the original Axiom book from 1992 with title %
``\emph{AXIOM---The Scientific Computation System}'' by Jenks and
Sutor in its adaptation for the \Language{} fork of the Axiom source
code.

It counts as the official version for the \Language{} project.

The original sources (including the text and scripts) for this book
where released in 2002 under the modified BSD license that is found in
the file
\href{https://github.com/fricas/fricas/blob/master/LICENSE.txt}{LICENSE.txt}
in the source code repository of \Language{}.

\InputIfFileExists{info.tex}{}{}

\section*{Warning}
Since this book is a moving target, each part of its generation is
automated. In particular, the output of \Language{} commands
(including their typesetting in \LaTeX{}) is automated. However, that
sometimes means that you see boxes with output that does not fit the
width of the page. We use the breqn \LaTeX{} package to break
equations appropriately. Unfortunately, that does not always work
nicely.

If you find other inaccuracies, please report them to the mailing list
\href{mailto:fricas-devel@googlegroups.com}{fricas-devel@googlegroups.com}.

\section*{Authors}

The book is an effort of quite a number of people.
%
The original Axiom book lists the following people as authors:
Manuel Bronstein,
William H. Burge,
Timothy P. Daly,
Patrizia Gianni,
Johannes Grabmeier,
Scott C. Morrison,
Jonathan M. Steinbach,
Robert S. Sutor,
Barry M. Trager,
Stephen M. Watt,
Richard D. Jenks,
Clifton J. Williamson.

Since \Language{} has been forked from the Axiom project, there have
been a number of contributions and corrections by Martin Baker,
Riccardo Guida, Waldek Hebisch, Ralf Hemmecke.


%\input{dedicate}
%\input{foreword}
\clearpage

\tableofcontents
%\input{bios}
%
\pagenumbering{arabic}
\setcounter{chapter}{-1}
\input{ug00}
\input{tecintro}  % This ha no chapter number.

\part{Basic Features of \Language{}}
%
\input{ug01}
\input{ug02}
\input{ug03}
\input{ug04}
\input{ug05}
\input{ug06}

%--rhx: TODO:
% Unfortunately ug07 uses \item[\spadfun{foo}] which doesn't work
% since \spadfun has a verbatim-like argument. Fortunately, that only
% happens inside the description environment. So we simply redefine it
% for this chapter. Better rewrite the text in ug07.
\begingroup
\let\spaddescriptionsave\description
\def\description{\def\spadfun{\textspadfun}\spaddescriptionsave}
\input{ug07}
\endgroup

\part{Advanced Problem Solving and Examples}
%
\input{ug08}
\input{ug09}

\part{Advanced Programming in \Language{}}
%
\input{ug10}
\input{ug11}
\input{ug12}
\input{ug13}
\input{ug14}
\input{ug15}
%
\appendix

\input{ug16}
%\input{ug17}
%\input{ug18}
%\input{ug19}
%\input{ug20}
\input{ug21}
%
\printindex
\end{document}
