\documentclass{article}
\usepackage{verbatim}
\usepackage{srcltx}
\usepackage{hyperref}
\newcommand{\variable}{\begingroup \urlstyle{tt}\Url}
\newcommand{\file}{\begingroup \urlstyle{tt}\Url}
\newcommand{\directory}{\begingroup \urlstyle{tt}\Url}
\newcommand{\code}{\begingroup \urlstyle{tt}\Url}
\newcommand{\defineterm}[1]{\textbf{#1}}
\newcommand{\useterm}[1]{#1}
\newcommand{\xAldor}{\textsc{Aldor}}
\newcommand{\xFriCAS}{\textsc{FriCAS}}

\begin{document}
\title{Comments Building the \xAldor{}-\xFriCAS{} Interface}
\author{Ralf Hemmecke}
\date{12-Aug-2008}
\maketitle
\begin{abstract}
  We give an overview of the way the \xAldor{}-\xFriCAS{} connection is
  built. Unfortunately, due to lack of proper knowledge of the
  internals of the \xAldor{} compiler and \xFriCAS{}, some parts here
  consist of pure guessing.
\end{abstract}
\tableofcontents

%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
\section{Overview}
%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%

Due to history, the code that is needed to build the
\xAldor{}-\xFriCAS{} interface is split into two kinds of files.
\begin{enumerate}
  \item The code available in the \xFriCAS{} repository, and
  \item code that lives in the \xAldor{} repository.
\end{enumerate}

Pre-existing code from \xFriCAS{} that is used in the process lives in
\file{src/interp/ax.boot}. After having built \file{libfricas.al}, the
file \file{src/interp/foam_l.lsp} is also needed.
For loading the \xAldor{} related stuff into \xFriCAS{} also
\file{src/interp/daase.lisp} is of importance.

We base the build process on previous work
by Peter Broadbery, and add the relevant files to the directory
\directory{src/aldor}.

One part of the \xAldor{} code that is needed for building the library
\file{libfricas.al} lives in the subdirectory \directory{libax0} in the
\xAldor{} repository. Only the file \file{lang.as} had to be taken
from the subdirectory \directory{axllib}.

The \xAldor{} code is split into several parts.
\begin{enumerate}
\item The \file{fricas.as} is globally needed as an include file for
  the build process.

\item Code that is needed as the very base of \file{libfricas.al}. It
  basically enables the \xAldor{} compiler to compile anything usefull
  at all.

\item The files \file{axlit.as} and \file{axextend.as} contain code
  that is intended to extend some domains from the \xFriCAS{} library.
\end{enumerate}



%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
\section{Basic Thoughts}
%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%

There are two difficulties in building the \xAldor{}-\xFriCAS{}
connection:
\begin{enumerate}
\item the \xFriCAS{} library contains a number of domains that depend
  on each other, and
\item the files \file{axlit.as} and \file{axextend.as} extend some
  \xFriCAS{} domains. On the other hand the extended functionality of
  \file{axlit.as} and \file{axextend.as} is needed for other domains.
\end{enumerate}

To solve the second problem, we must compile \file{axlit.as} and
\file{axextend.as} somewhere in the middle of the \xFriCAS{} library.

Let us first give the main idea to solve both problems.
%
The full list of types of the \xFriCAS{} library is searched for
cliques, i.e., domains that depend on each other. For each clique one
\texttt{.ap} file will be constructed and this will be fed to the
\xAldor{} compiler for inclusion into \file{libfricas.al}.
%
Note that each clique will be named after its lexicographically
smallest member where the names are the abbreviated forms of \xFriCAS{}
constructors. Thus, a file like, for example, \file{COMPCAT.ap}
will also contain definitions for the constructor \texttt{COMPLEX}.

Among those cliques a compilation order is determined. Then the
compilation of \file{axlit.as} and \file{axextend.as} is introduced as
early as possible in this order, i.e., just after all cliques on which
they depend are already compiled.

It might be that without the following trick it is impossible to find
a place in the list of cliques, so that \file{axlit.as} and
\file{axextend.as} can be compiled on their own. That is just a guess
and has not been experimentally confirmed.

There is a trick to reduce the size of the cliques. If a domain or
category $A$ mentions another domain $B$ then it is not always
necessary that all knowledge of $B$ must be available to compile (the
interface of) $A$. See \file{src/interp/ax.boot.tex} for more detail
of how the SPAD code of $A$ is turned into a dummy domain definition
(or basically an interface) of $A$ that will go into
\file{libfricas.al}.

For example, if $A$ only exports a function
\begin{verbatim}
foo: B -> %
\end{verbatim}
then it is sufficient if an initial domain
\begin{verbatim}
B: with == add;
\end{verbatim}
has been compiled before $A$ and the actual (full) definition of $B$
is turned into
\begin{verbatim}
extend B: CatB == (add pretend CatB) add;
\end{verbatim}}
and will be compiled after $A$. (\code{CatB} above stands for the full
list of exports taken from the (compiled) SPAD code.)

One can go even further and just provide
\begin{verbatim}
export B: with;
\end{verbatim}
Such an initial list was handcrafted by Peter Broadbery with a few
additions by Ralf Hemmecke and is reflected in
\file{src/aldor/initlist.as}.



%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
\section{What are the Interface Files?}
%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%

The code that will go into \file{libfricas.al} can be classified as
follows.

\begin{enumerate}
\item Code that is needed as the very base of \file{libfricas.al}. This
  part is completely independent of \xFriCAS{}. It consists of the
  files \file{lang.as}, \file{minimach.as}, and \file{boolean0.as}.

\item From the file \file{stub.as} we will extract the definition of
  the category \code{SubsetCategory}, which is a builtin type in SPAD.
  That definition will be written to a file \file{SUBSETC.as} and as
  that made available in \file{libfricas.al}.

\item The file \file{attrib.as} is not taken from the \xAldor{}
  repository, but rather generated from information in \xFriCAS{}. It
  contains definitions of categories, one for each attribute available
  in \xFriCAS{}.

\item Some initial domains, that can be found in \file{initlist.as}.

\item The \texttt{.ap} form of \xFriCAS{} cliques which are
  prerequisites of \file{axlit.as} and \file{axextend.as}. We call
  those types \defineterm{lower types}.

\item The files \file{axlit.as} and \file{axextend.as}.

\item The \texttt{.ap} form of \xFriCAS{} cliques which are not
  prerequisites of \file{axlit.as} and \file{axextend.as}. We call
  those types \defineterm{upper types}.
\end{enumerate}

Note that the \texttt{.ap} files will be generated by
\file{gendepap.lsp} with the help of the function
\verb'|makeAxExportForm|' from \file{src/interp/ax.boot}.


%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
\section{Building the \xAldor{}-\xFriCAS{} Interaction Files}
%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
For the interaction we must create \file{libfricas.al}.

ToDo: From \file{src_aldor2} (see
\url{fricas-wiki.math.uni.wroc.pl/uploads/src_aldor2.tgz}) it
seems that more files (in particular some \texttt{.lsp} files must
then be created.

%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
\section{Building libfricas.al}
%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%

The process of building \file{libfricas.al} happens in two stages.
\begin{enumerate}
\item The commands in \file{Makefile.in} are mainly responsible for
  generating the file \file{cliques.mk}.
%
  The latter is a partial Makefile and basically contains for each
  clique its dependencies.

\item The actual creation of \file{libfricas.al} is done via the
  \file{Makefile2.in} using \file{cliques.mk}.
\end{enumerate}



%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
\subsection{Computing Dependencies}
%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%

The process of creating the file \file{cliques.mk} is controlled by
\file{Makefile.in} with help of \file{gendepap.lsp} and
\file{src/interp/ax.boot}.

It is done with the following steps:
\begin{enumerate}
\item Create the file \file{domains.mk} containing all domains,
  packages, and categories of the \xFriCAS{} library.

\item Get some non-free files from an \xAldor{} repository (if they do
  not yet exist in the build directory).

\item Generate the files \file{SUBSETC.as} from \file{stub.as} and
  \file{attrib.as} from the actual attribute list inside \xFriCAS{}.

\item From \file{initlist.as} create \file{initlist.ap}. The latter
  file is used in \file{gendepap.lsp} to set the variable
  \verb'$extendedDomains'. %$
  These extended domains should not be confused with the domains that
  are extended via \file{axlit.as} and \file{axextend.as}. We would
  rather like to call this variables \verb'$initDomains', %$
  but for this change we would have to modify
  \file{src/interp/ax.boot}.

\item Compute dependency lists.
  \begin{enumerate}
  \item For each export $T$ in \file{initlist.as} create a separate
    file of the form \file{init_A.ap} and dependencies of that file
    into \file{init_A.dep} where $A$ is the abbreviated form of $T$.

  \item For each type $T$ from \file{domains.mk} (except some
    blacklisted ones) compute \file{A.dep} where $A$ is the
    abbreviated form of $T$.

  \item For each file \texttt{F.as} (see variable
    \variable{aldor_srcs} in \file{Makefile.in}) compute its
    dependencies and write them to \file{F.dep}.
  \end{enumerate}

  Note that the algorithm in \file{src/aldor/gendepap.lsp} to compute
  dependencies takes information from \file{initlist.as} into account.
%
  It generates a dependency on \verb'init_A' instead of \texttt{A} if
  an initial definition of the form
\begin{verbatim}
export A: with;
\end{verbatim}
  (or similarly simple) is sufficient to compile the \texttt{.ap} file
  in question.

\item Create the file \file{libfricas.lst} that will be used by
  \file{cliques.as} to know about all files to consider.

\item Call \file{cliques} on \file{libfricas.lst} and all the
  dependency files and thus compute \file{cliques.mk}.
\end{enumerate}






%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
\subsection{Compiling the Library}
%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%



%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
\section{The Extension Files}
%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%

Looking at the files \file{axlit.as} and \file{axextend.as} more
closely, reveals that they should make the \xAldor{} language defined
types like \code{Literal} and \code{Generator} known.
%
Furthermore, some code deals with extending \xFriCAS{} types to make
them behave according to the Aldor User Guide, for example
\code{Tuple}.
%
Another part (according to Peter Broadbery
\url{http://groups.google.com/group/fricas-devel/browse_thread/thread/8db28eab0cb536bd/c8bafffdd6bd8fc3#c8bafffdd6bd8fc3})
is for helping the \xAldor{} compiler with inlining certain
statements.

Unfortunately, the different purposes are not clear from the
(non-existing) documentation. It also seems worth to mention that some
domains in \file{axlit.as} and \file{axextend.as} explicitly assume a
certain representation. These are \code{List}, \code{Tuple},
\code{Vector}, \code{Matrix}, \code{Segment}, and
\code{UniversalSegment}.

%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
\subsection{Which Domains Need more Features?}
\label{sec:ExtendedDomains}
%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
Domains extended in \file{axlit.as} and \file{axextend.as}.
  \begin{description}
    \item \file{axlit.as}:
      \begin{quote}
        \code{String},
        \code{Symbol},
        \code{SingleInteger},
        \code{Integer},
        \code{NonNegativeInteger},
        \code{PositiveInteger},
        \code{DoubleFloat},
        \code{Float},
        \code{Tuple}, and
        \code{List}
      \end{quote}
    \item \file{axlit.as}:
      \begin{quote}
        \code{Symbol},
        \code{SingleInteger},
        \code{Integer},
        \code{NonNegativeInteger},
        \code{PositiveInteger},
        \code{DoubleFloat},
        \code{Float},
        \code{Tuple},
        \code{List},
        \code{Vector},
        \code{Matrix},
        \code{Segment}, and
        \code{UniversalSegment}
      \end{quote}
  \end{description}



%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
\subsection{Known Issues}
\label{sec:KnownIssues}
%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
If \file{libfricas.al} is build incrementally without intermediate
libraries of the form \file{libfricas_D.al}, its size is about
27324110 bytes which is about 3 times as big as the 9320874 that is
build with srcaldor2 by Peter Broadbery.



\end{document}
